Refinements for Open Automata

### 2023-09-06 2021-Refinement/paper/SEFM-Extend-RR.tex
version modif.

%%\usepackage[frenchb]{babel} % optionnel
\RRNo{7003}
%% \RTNo{0703}
%%
%% date de publication du rapport
\RRdate{August 2023}
%%
%% Cas d'une version deux
%% \RRversion{2}
%% date de publication de la version 2
%% \RRdater{November 2008}
%%
\RRauthor{% les auteurs
 % Premier auteur, avec une note
Rab\'ea Ameur-Boulifa \thanks{LTCI, T\'el\'ecom Paris, Institut Polytechnique de Paris}%
  % note partag\'ee (optionnelle)
 % \and entre chaque auteur s'il y en a plusieurs
  \and
Quentin Corradi \thanks[sfn]{Universit\'e Lyon, EnsL, UCBL, CNRS, Inria,  LIP}%
 % r\'ef\'erence \`a la note partag\'ee
 % liste longue pour tests de mise en page
\and Ludovic Henrio \thanksref{sfn}   \and Eric Madelaine \thanks{INRIA Sophia Antipolis M\'edit\'erann\'ee, UCA}}
%% Ceci apparait sur chaque page paire.
\authorhead{Rab\'ea Ameur-Boulifa et al.}
%% titre francais long
\RRtitle{Raffinements des Automates Ouverts}
%% English title
\RRetitle{Refinements for Open Automata\\ (Extended Version)}
%%
\titlehead{Refinements of Open Automata}
%%
%\RRnote{This is a note}
%\RRnote{This is a second note}
%%
\RRresume{
Vérifier la correction d'un programme est une tâche complexe souvent réalisée de manière progressive en établissant une relation entre les comportements de deux programmes.  La relation de simulation, qui met en corespondance   une spécification et une implémentation, ou un programme et sa version raffinée, permet d'assurer la correction. Cet article introduit et discute d'une relation entre des programmes représentés sous la forme d'\emph{automates ouverts}, un modèle comportemental symbolique pour les systèmes de communication. Établir des relations d'équivalence et de raffinement sur les automates ouverts permet une approche compositionnelle pour vérifier leur comportement. L'aspect compositionnel des automates ouverts provient des \emph{trous}, qui représentent des éléments du contexte. La composition de deux automates consiste à instancier un trou d'un automate avec l'autre automate. Dans cet article, nous définissons une relation de raffinement entre les automates ouverts qui peuvent avoir ou non les mêmes trous, sur le modèle des simulations. Nous montrons ensuite dans quelles conditions la composition d'automates ouverts préserve cette relation.

}
\RRabstract{
Establishing equivalence and refinement relations between programs is an important mean for 
verifying their correctness. By establishing that the
behaviours of a modified program simulate those of the source one, simulation relations formalise the desired relationship between a specification and an implementation,  two equivalent implementations, or a program and its optimised implementation.
This article discusses a notion of simulation between \emph{open automata}, which are symbolic behavioural models for communicating systems. 
Open automata may have \emph{holes} modelling elements of their
context, and can be composed by instantiation of the holes. This allows for a compositional approach for verification of their behaviour. We define a simulation between open automata that may or may not have the same holes, and 
show under which conditions these refinements are preserved by composition of open automata.
 }
%%
\RRmotcle{Syst\`emes de transitions \'etiquet\'ees, Simulation, Raffinement, Composition}
\RRkeyword{
Labelled transition systems, Simulation, Refinement, Composition}
%%
%% \RRprojet{Apics}  % cas d'un seul projet
\RRprojets{LTCI and LIP and KAIROS}
%%
%% \URLorraine % pour ceux qui sont \`a l'est
%% \URRennes  % pour ceux qui sont \`a l'ouest
%% \URRhoneAlpes % pour ceux qui sont dans les montagnes
%% \URRocq % pour ceux qui sont au centre de la France
%% \URFuturs % pour ceux qui sont dans le virtuel
%% \URSophia % pour ceux qui sont au Sud.
%%
%% \RCBordeaux % centre de recherche Bordeaux - Sud Ouest
%% \RCLille % centre de recherche Lille Nord Europe
%% \RCParis % Paris Rocquencourt
%% \RCSaclay % Saclay \^Ile de France
%% \RCGrenoble % Grenoble - Rh\^one-Alpes
%% \RCNancy % Nancy - Grand Est
%% \RCRennes % Rennes - Bretagne Atlantique
\RCSophia % Sophia Antipolis M\'editerran\'ee

@@

\proof
Let us denote by $A_{13}$ (resp. $A_{23}$) the OA resulting from $A_1\subst{A_3}{k}$ (resp. $A_2\subst{A_3}{k}$),  to prove the theorem it is sufficient to prove that there exists a relation between states of the two OA that satisfies the conditions of  Definition~\ref{Def:OA-Refinement}.

We denote $A_1=  \OA{S_1}{s_{01}}{J_1}{V_1}{\sigma_{01}}{T_1}$ and \(A_2 \defobject \OA{S_2}{s_{02}}{J_2}{V_2}{\sigma_{02}}{T_2}\) and $A_3 = \OA{S_3}{s_{03}}{J_3}{V_3}{\sigma_{03}}{T_3}$. The proof requires  to rename the variables of one instance of the two $A_3$ automata to avoid clashes in variable names (this is required by the definition of refinement). In practice we will use superscripts ${}^1$ and ${}^2$ to distinguish elements of the two instances of $A_3$.

@@

We want to prove that $\mpar{\mathcal{R}', H\uplus J_3 \setminus\{k\}}$ is a hole-tracking simulation of $A_{13}$ and $A_{23}$. In the following we denote $H'=H\cup J_3 \setminus\{k\}$.
\begin{enumerate}
\item First, we have to prove the relation for initial states:
\[\sigma_{013} \uplus \sigma_{023} \vdash \mathcal{R}'\mpar{s_{013}, s_{023}}\]
with $\sigma_{013} = \sigma_{01} \uplus \sigma_{03}^1$, $\sigma_{023} = \sigma_{02} \uplus \sigma_{03}^2$, $s_{013}=(s_{01},s_{03}^1)$, and $s_{023}=(s_{02},s_{02}^2)$.\\
By using the fact that $\mathcal{R}$ relates initial configurations of  $A_1$ and $A_2$, we have:
$\mpar{\sigma_{01} \uplus \sigma_{02} \vdash \mathcal{R}\mpar{s_{01}, s_{02}}}$.  
%and based on the  fact that initial states are reachable $\sigma_{013} \vdash \reach{A_{13}}\mpar{s_{013}}$ we have 
%\[ \mpar{\sigma_{01} \uplus \sigma_{02} \vdash R\mpar{s_{01}, s_{02}}} \wedge 
%\mpar{\sigma_{013} \vdash \reach{A_{13}}\mpar{s_{013}}}\]

Considering that initial valuations $\sigma_{03}^1$ and $\sigma_{03}^2$ associate the same values to the ``same'' variables modulo renaming, so the following holds:\\ $\left(\,\,{\displaystyle \bigwedge_\subbox{v_3\in V_3} v^1_3=v^2_3\wedge s^1_3 = s^2_3}\right)\psubst{\sigma_{03}^1 \uplus \sigma_{03}^2 }$.

@@

\end{enumerate}

\end{document}
\endinput
%%
%% End of file `squelette-rr.tex'.

>>>>

%%\usepackage[frenchb]{babel} % optionnel
\RRNo{9517}
%% \RTNo{0703}
%%
%% date de publication du rapport
\RRdate{August 2023}
%%
%% Cas d'une version deux
%% \RRversion{2}
%% date de publication de la version 2
%% \RRdater{November 2008}
%%
\RRauthor{% les auteurs
 % Premier auteur, avec une note
Rab\'ea Ameur-Boulifa \thanks{LTCI, T\'el\'ecom Paris, Institut Polytechnique de Paris}%
  % note partag\'ee (optionnelle)
 % \and entre chaque auteur s'il y en a plusieurs
  \and
Quentin Corradi \thanks[sfn]{Universit\'e Lyon, EnsL, UCBL, CNRS, Inria,  LIP}%
 % r\'ef\'erence \`a la note partag\'ee
 % liste longue pour tests de mise en page
\and Ludovic Henrio \thanksref{sfn}   \and Eric Madelaine \thanks{INRIA Sophia Antipolis M\'edit\'erann\'ee, UCA}}
%% Ceci apparait sur chaque page paire.
\authorhead{Rab\'ea Ameur-Boulifa et al.}
%% titre francais long
\RRtitle{Raffinements des automates ouverts}
%% English title
\RRetitle{Refinements for Open Automata\\ (Extended Version)}
%%
\titlehead{Refinements of Open Automata}
%%
%\RRnote{This is a note}
%\RRnote{This is a second note}
%%
\RRresume{
Vérifier la correction d'un programme est une tâche complexe souvent réalisée de manière progressive en établissant une relation entre les comportements de deux programmes.  La relation de simulation, qui met en correspondance   une spécification et une implémentation, ou un programme et sa version raffinée, permet d'assurer la correction. Cet article introduit et discute d'une relation entre des programmes représentés sous la forme d'\emph{automates ouverts}, un modèle comportemental symbolique pour les systèmes de communication. Établir des relations d'équivalence et de raffinement sur les automates ouverts permet une approche compositionnelle pour vérifier leur comportement. L'aspect compositionnel des automates ouverts provient des \emph{trous}, qui représentent des éléments du contexte. La composition de deux automates consiste à instancier un trou d'un automate avec l'autre automate. Dans cet article, nous définissons une relation de raffinement entre les automates ouverts qui peuvent avoir ou non les mêmes trous, sur le modèle des simulations. Nous montrons ensuite dans quelles conditions la composition d'automates ouverts préserve cette relation.

}
\RRabstract{
Establishing equivalence and refinement relations between programs is an important mean for 
verifying their correctness. By establishing that the
behaviours of a modified program simulate those of the source one, simulation relations formalise the desired relationship between a specification and an implementation,  two equivalent implementations, or a program and its optimised implementation.
This article discusses a notion of simulation between \emph{open automata}, which are symbolic behavioural models for communicating systems. 
Open automata may have \emph{holes} modelling elements of their
context, and can be composed by instantiation of the holes. This allows for a compositional approach for verification of their behaviour. We define a simulation between open automata that may or may not have the same holes, and 
show under which conditions these refinements are preserved by composition of open automata.
 }
%%
\RRmotcle{Syst\`emes de transitions \'etiquet\'ees, Simulation, Raffinement, Composition}
\RRkeyword{
Labelled transition systems, Simulation, Refinement, Composition}
%%
%% \RRprojet{Apics}  % cas d'un seul projet
\RRprojets{CASH and KAIROS}
%%
%% \URLorraine % pour ceux qui sont \`a l'est
%% \URRennes  % pour ceux qui sont \`a l'ouest 
\URRhoneAlpes % pour ceux qui sont dans les montagnes
%% \URRocq % pour ceux qui sont au centre de la France
%% \URFuturs % pour ceux qui sont dans le virtuel
% \URSophia % pour ceux qui sont au Sud.
%%
%% \RCBordeaux % centre de recherche Bordeaux - Sud Ouest
%% \RCLille % centre de recherche Lille Nord Europe
%% \RCParis % Paris Rocquencourt
%% \RCSaclay % Saclay \^Ile de France
%% \RCGrenoble % Grenoble - Rh\^one-Alpes
%% \RCNancy % Nancy - Grand Est
%% \RCRennes % Rennes - Bretagne Atlantique
%\RCSophia % Sophia Antipolis M\'editerran\'ee

@@

\proof
Let us denote by $A_{13}$ (resp. $A_{23}$) the OA resulting from $A_1\subst{A_3}{k}$ (resp. $A_2\subst{A_3}{k}$),  to prove the theorem it is sufficient to prove that there exists a relation between states of the two OA that satisfies the conditions of  Definition~\ref{Def:OA-Refinement}.  We denote $A_1=  \OA{S_1}{s_{01}}{J_1}{V_1}{\sigma_{01}}{T_1}$, \(A_2 = \OA{S_2}{s_{02}}{J_2}{V_2}{\sigma_{02}}{T_2}\) and $A_3 = \OA{S_3}{s_{03}}{J_3}{V_3}{\sigma_{03}}{T_3}$. The proof requires  to rename the variables of one instance of the two $A_3$ automata to avoid clashes in variable names (this is required by the definition of refinement). In practice we will use superscripts ${}^1$ and ${}^2$ to distinguish elements of the two instances of $A_3$.

@@

We want to prove that $\mpar{\mathcal{R}', H\uplus J_3 \setminus\{k\}}$ is a hole-tracking simulation of $A_{13}$ and $A_{23}$. In the following we denote $H'=H\cup J_3 \setminus\{k\}$.
\begin{enumerate}
\item First, we have to prove the relation for initial states:
\[\sigma_{013} \uplus \sigma_{023} \vdash \mathcal{R}'\mpar{s_{013}, s_{023}}\]
with $\sigma_{013} = \sigma_{01} \uplus \sigma_{03}^1$, $\sigma_{023} = \sigma_{02} \uplus \sigma_{03}^2$, $s_{013}=(s_{01},s_{03}^1)$, and $s_{023}=(s_{02},s_{02}^2)$.\\
By using the fact that $\mathcal{R}$ relates initial configurations of  $A_1$ and $A_2$, we have:
\[\mpar{\sigma_{01} \uplus \sigma_{02} \vdash \mathcal{R}\mpar{s_{01}, s_{02}}}\]  
%and based on the  fact that initial states are reachable $\sigma_{013} \vdash \reach{A_{13}}\mpar{s_{013}}$ we have 
%\[ \mpar{\sigma_{01} \uplus \sigma_{02} \vdash R\mpar{s_{01}, s_{02}}} \wedge 
%\mpar{\sigma_{013} \vdash \reach{A_{13}}\mpar{s_{013}}}\]
Considering that initial valuations $\sigma_{03}^1$ and $\sigma_{03}^2$ associate the same values to the ``same'' variables modulo renaming, so the following holds:\\ $\left(\,\,{\displaystyle \bigwedge_\subbox{v_3\in V_3} v^1_3=v^2_3\wedge s^1_3 = s^2_3}\right)\psubst{\sigma_{03}^1 \uplus \sigma_{03}^2 }$.

@@

\end{enumerate}
\newpage
\tableofcontents
\end{document}
\endinput
%%
%% End of file `squelette-rr.tex'.
